ECCCos from the Black Box:\\ Faithful Explanations through\\ Energy-Constrained Conformal Counterfactuals

### 2023-09-20 paper/appendix.tex
shoot me
more work on the appendix

Since we were not able to identify any existing open-source software for Energy-Based Modelling that would be flexible enough to cater to our needs, we have developed a \texttt{Julia} package from scratch. The package has been open-sourced, but to avoid compromising the double-blind review process, we refrain from providing more information at this stage. In our development we have heavily drawn on the existing literature:~\citet{du2020implicit} describe best practices for using EBM for generative modelling;~\citet{grathwohl2020your} explain how EBM can be used to train classifiers jointly for the discriminative and generative tasks. We have used the same package for training and inference, but there are some important differences between the two cases that are worth highlighting here.

@@

To train our Joint Energy Models we broadly follow the approach outlined in~\citet{grathwohl2020your}. These models are trained to optimize a hybrid objective that involves a standard classification loss component $L_{\text{clf}}(\theta)=-\log p_{\theta}(\mathbf{y}|\mathbf{x})$ (e.g. cross-entropy loss) as well as a generative loss component $L_{\text{gen}}(\theta)=-\log p_{\theta}(\mathbf{x})$.

To draw samples from $p_{\theta}(\mathbf{x})$, we rely exclusively on the conditional sampling approach described in~\citet{grathwohl2020your} for both training and inference: we first draw $\mathbf{y}\sim p(\mathbf{y})$ and then sample $\mathbf{x} \sim p_{\theta}(\mathbf{x}|\mathbf{y})$~\citep{grathwohl2020your} via Equation~\ref{eq:sgld} with energy $\mathcal{E}(\mathbf{x}|\mathbf{y})=\mu_{\theta}(\mathbf{x})[\mathbf{y}]$ where $\mu_{\theta}: \mathcal{X} \mapsto \mathbb{R}^K$ returns the linear predictions (logits) of our classifier $M_{\theta}$. While our package also supports unconditional sampling, we found conditional sampling to work well. It is also well aligned with CE, since in this context we are interested in conditioning on the target class.

@@

\begin{equation}\label{eq:biased-sgld}
  \begin{aligned}
    \hat{\mathbf{x}}_{j+1} &\leftarrow \hat{\mathbf{x}}_j - \frac{\epsilon}{2} \mathcal{E}(\hat{\mathbf{x}}_j|\mathbf{y}^+) + \sigma \mathbf{r}_j, && j=1,...,J
  \end{aligned}
\end{equation}

@@

It is important to realise that sampling is done during each training epoch, which makes training Joint Energy Models significantly harder than conventional neural classifiers. In each epoch the generated (batch of) sample(s) $\hat{\mathbf{x}}_{J}$ is used as part of the generative loss component, which compares its energy to that of observed samples $\mathbf{x}$: $L_{\text{gen}}(\theta)=\mu_{\theta}(\mathbf{x})[\mathbf{y}]-\mu_{\theta}(\hat{\mathbf{x}}_{J})[\mathbf{y}]$. Our full training objective can be summarized as follows,

@@

where $L_{\text{reg}}(\theta)$ is a Ridge penalty (L2 norm) that regularises energy magnitudes for both observed and generated samples~\citep{du2020implicit}. We have used varying degrees of regularization depending on the dataset ($\lambda$ in Table~\ref{tab:ebmparams}).

@@

\begin{equation} \label{eq:eccco-app}
  \begin{aligned}
  \mathbf{Z}^\prime &= \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^L} \{  {\text{yloss}(M_{\theta}(f(\mathbf{Z}^\prime)),\mathbf{y}^+)}+ \lambda_{1} {\text{dist}(f(\mathbf{Z}^\prime),\mathbf{x}) } \\
  &+ \lambda_2 \Delta\mathcal{E}(\mathbf{Z}^\prime,\widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}) + \lambda_3 \Omega(C_{\theta}(f(\mathbf{Z}^\prime);\alpha)) \} 
  \end{aligned} 
\end{equation}

@@

since $\Delta\mathcal{E}(\cdot)$ is equivalent to the generative loss function $L_{\text{gen}}(\cdot)$. In fact, this is also true for $\lambda L_{\text{reg}}(\theta)\ne0$ since we use the Ridge penalty $L_{\text{reg}}(\theta)$ in the counterfactual search just like we do in joint-energy training. This detail was omitted from the body of the paper for the sake of simplicity.

Aside from the additional penalties in Equation~\ref{eq:eccco-app}, the only key difference between our counterfactual search objective and the joint-energy training objective is the parameter that is being optimized. In joint-energy training we optimize the objective with respect to the network weights $\theta$:

\begin{equation}\label{eq:jem-grad}
  \begin{aligned}
    \nabla_{\theta} L_{\text{JEM}}(\theta) &=\nabla_{\theta} L_{\text{clf}}(\theta) + \nabla_{\theta}L_{\text{gen}}(\theta) + \lambda \nabla_{\theta} L_{\text{reg}}(\theta) 
  \end{aligned}
\end{equation}

we take these parameters as fixed in our counterfactual search.

>>>>

Since we were not able to identify any existing open-source software for Energy-Based Modelling that would be flexible enough to cater to our needs, we have developed a \texttt{Julia} package from scratch. The package has been open-sourced, but to avoid compromising the double-blind review process, we refrain from providing more information at this stage. In our development we have heavily drawn on the existing literature:~\citet{du2019implicit} describe best practices for using EBM for generative modelling;~\citet{grathwohl2020your} explain how EBM can be used to train classifiers jointly for the discriminative and generative tasks. We have used the same package for training and inference, but there are some important differences between the two cases that are worth highlighting here.

@@

To train our Joint Energy Models we broadly follow the approach outlined in~\citet{grathwohl2020your}. Formally, JEMs are defined by the following joint distribution:

\begin{equation}
  \begin{aligned}
    \log p\_{\theta}(\mathbf{x},\mathbf{y}) &= \log p_{\theta}(\mathbf{y}|\mathbf{x}) + \log p_{\theta}(\mathbf{x})
  \end{aligned}
\end{equation}

Training therefore involves a standard classification loss component $L_{\text{clf}}(\theta)=-\log p_{\theta}(\mathbf{y}|\mathbf{x})$ (e.g. cross-entropy loss) as well as a generative loss component $L_{\text{gen}}(\theta)=-\log p_{\theta}(\mathbf{x})$. Analogous to how we defined the conditional distribution over inputs in Definition~\ref{def:faithful}, $p_{\theta}(\mathbf{x})$ denotes the unconditional distribution over inputs. The model gradient of this component of the loss function can be expressed as follows:

\begin{equation}\label{eq:gen-true}
  \begin{aligned}
    \nabla_{\theta}L_{\text{gen}}(\theta)&=-\nabla_{\theta}\log p\_{\theta}(\mathbf{x})=-\left(\mathbb{E}_{p(\mathbf{x})} \left\{  \nabla_{\theta} \mathcal{E}_{\theta}(\mathbf{x}) \right\} - \mathbb{E}_{p_{\theta}(\mathbf{x})} \left\{  \nabla_{\theta} \mathcal{E}_{\theta}(\mathbf{x}) \right\} \right)
  \end{aligned}
\end{equation}

To draw samples from $p_{\theta}(\mathbf{x})$, we rely exclusively on the conditional sampling approach described in~\citet{grathwohl2020your} for both training and inference: we first draw $\mathbf{y}\sim p(\mathbf{y})$ and then sample $\mathbf{x} \sim p_{\theta}(\mathbf{x}|\mathbf{y})$~\citep{grathwohl2020your} via Equation~\ref{eq:sgld} with energy $\mathcal{E}_{\theta}(\mathbf{x}|\mathbf{y})=\mu_{\theta}(\mathbf{x})[\mathbf{y}]$ where $\mu_{\theta}: \mathcal{X} \mapsto \mathbb{R}^K$ returns the linear predictions (logits) of our classifier $M_{\theta}$. While our package also supports unconditional sampling, we found conditional sampling to work well. It is also well aligned with CE, since in this context we are interested in conditioning on the target class.

@@

\begin{equation}\label{eq:biased-sgld}
  \begin{aligned}
    \hat{\mathbf{x}}_{j+1} &\leftarrow \hat{\mathbf{x}}_j - \frac{\epsilon}{2} \mathcal{E}_{\theta}(\hat{\mathbf{x}}_j|\mathbf{y}^+) + \sigma \mathbf{r}_j, && j=1,...,J
  \end{aligned}
\end{equation}

@@

It is important to realise that sampling is done during each training epoch, which makes training Joint Energy Models significantly harder than conventional neural classifiers. In each epoch the generated (batch of) sample(s) $\hat{\mathbf{x}}_{J}$ is used as part of the generative loss component, which compares its energy to that of observed samples $\mathbf{x}$:

\begin{equation}\label{eq:gen-loss}
  \begin{aligned}
    L_{\text{gen}}(\theta)&\approx\mu_{\theta}(\mathbf{x})[\mathbf{y}]-\mu_{\theta}(\hat{\mathbf{x}}_{J})[\mathbf{y}]
  \end{aligned}
\end{equation}

Our full training objective can be summarized as follows,

@@

where $L_{\text{reg}}(\theta)$ is a Ridge penalty (L2 norm) that regularises energy magnitudes for both observed and generated samples~\citep{du2019implicit}. We have used varying degrees of regularization depending on the dataset ($\lambda$ in Table~\ref{tab:ebmparams}).

@@

\begin{equation} \label{eq:eccco-app}
  \begin{aligned}
  \mathbf{Z}^\prime &= \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^L} \{  {\text{yloss}(M_{\theta}(f(\mathbf{Z}^\prime)),\mathbf{y}^+)}+ \lambda_{1} {\text{dist}(f(\mathbf{Z}^\prime),\mathbf{x}) } \\
  &+ \lambda_2 \Delta\mathcal{E}_{\theta}(\mathbf{Z}^\prime,\widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}) + \lambda_3 \Omega(C_{\theta}(f(\mathbf{Z}^\prime);\alpha)) \} 
  \end{aligned} 
\end{equation}

@@

since $\Delta\mathcal{E}_{\theta}(\cdot)$ is equivalent to the generative loss function $L_{\text{gen}}(\cdot)$. In fact, this is also true for $\lambda L_{\text{reg}}(\theta)\ne0$ since we use the Ridge penalty $L_{\text{reg}}(\theta)$ in the counterfactual search just like we do in joint-energy training. This detail was omitted from the body of the paper for the sake of simplicity.

Aside from the additional penalties in Equation~\ref{eq:eccco-app}, the only key difference between our counterfactual search objective and the joint-energy training objective is the parameter that is being optimized. In joint-energy training we optimize the objective with respect to the network weights $\theta$. Recall that $\mathcal{E}_{\theta}(\mathbf{x}|\mathbf{y})=\mu_{\theta}(\mathbf{x})[\mathbf{y}]$. Then the partial gradient with respect to the generative loss component can be expressed as follows:

\begin{equation}\label{eq:jem-grad}
  \begin{aligned}
    \nabla_{\theta}L_{\text{gen}}(\theta) &= \nabla_{\theta}\mu_{\theta}(\mathbf{x})[\mathbf{y}]- \nabla_{\theta}\mu_{\theta}(\hat{\mathbf{x}}_{J})[\mathbf{y}]
  \end{aligned}
\end{equation}

During the counterfactual search, we take the network parameters as fixed and instead optimize with respect to the counterfactual itself,

\begin{equation}\label{eq:ce-grad}
  \begin{aligned}
    \nabla_{\mathbf{x}}L_{\text{gen}}(\theta) &= \nabla_{\mathbf{x}}\mu_{\theta}(\mathbf{x})[\mathbf{y}]- \nabla_{\mathbf{x}}\mu_{\theta}(\hat{\mathbf{x}}_{J})[\mathbf{y}]
  \end{aligned}
\end{equation}

where we omit the notion of a latent search space to make the comparison easier. Intuitively, taking iterative gradient steps according to Equation~\ref{eq:ce-grad} has the effect of decreasing the energy of the counterfactual until it is in balance with the energy of conditional samples generated through SGLD.

### 2023-09-20 paper/body.tex
shoot me
more work on the appendix

\begin{equation}\label{eq:sgld}
  \begin{aligned}
    \mathbf{x}_{j+1} &\leftarrow \mathbf{x}_j - \frac{\epsilon_j^2}{2} \mathcal{E}(\mathbf{x}_j|\mathbf{y}^+) + \epsilon_j \mathbf{r}_j, && j=1,...,J
  \end{aligned}
\end{equation}

where $\mathbf{r}_j \sim \mathcal{N}(\mathbf{0},\mathbf{I})$ is the stochastic term and the step-size $\epsilon_j$ is typically polynomially decayed~\citep{welling2011bayesian}. The term $\mathcal{E}(\mathbf{x}_j|\mathbf{y}^+)$ denotes the model energy conditioned on the target class label $\mathbf{y}^+$ which we specify as the negative logit corresponding to the target class label $\mathbf{y}^{+}$. To allow for faster sampling, we follow the common practice of choosing the step-size $\epsilon_j$ and the standard deviation of $\mathbf{r}_j$ separately. While $\mathbf{x}_J$ is only guaranteed to distribute as $p_{\theta}(\mathbf{x}|\mathbf{y}^{+})$ if $\epsilon \rightarrow 0$ and $J \rightarrow \infty$, the bias introduced for a small finite $\epsilon$ is negligible in practice \citep{murphy2023probabilistic,grathwohl2020your}. Appendix~\ref{app:jem} provides additional implementation details for any tasks related to energy-based modelling.

@@

Our default choice for the $\text{dist}(\cdot)$ in both cases is the L1 Norm.

@@

\begin{equation} \label{eq:eccco}
  \begin{aligned}
  \mathbf{Z}^\prime &= \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^L} \{  {\text{yloss}(M_{\theta}(f(\mathbf{Z}^\prime)),\mathbf{y}^+)}+ \lambda_{1} {\text{dist}(f(\mathbf{Z}^\prime),\mathbf{x}) } \\
  &+ \lambda_2 \Delta\mathcal{E}(\mathbf{Z}^\prime,\widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}) + \lambda_3 \Omega(C_{\theta}(f(\mathbf{Z}^\prime);\alpha)) \} 
  \end{aligned} 
\end{equation}

@@

\begin{equation} \label{eq:energy-delta}
  \begin{aligned}
    \Delta\mathcal{E}&=\mathcal{E}(f(\mathbf{Z}^\prime)|\mathbf{y}^+)-\mathcal{E}(x|\mathbf{y}^+) &&& x \sim \widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}
  \end{aligned}
\end{equation}

>>>>

\begin{equation}\label{eq:sgld}
  \begin{aligned}
    \mathbf{x}_{j+1} &\leftarrow \mathbf{x}_j - \frac{\epsilon_j^2}{2} \mathcal{E}_{\theta}(\mathbf{x}_j|\mathbf{y}^+) + \epsilon_j \mathbf{r}_j, && j=1,...,J
  \end{aligned}
\end{equation}

where $\mathbf{r}_j \sim \mathcal{N}(\mathbf{0},\mathbf{I})$ is the stochastic term and the step-size $\epsilon_j$ is typically polynomially decayed~\citep{welling2011bayesian}. The term $\mathcal{E}_{\theta}(\mathbf{x}_j|\mathbf{y}^+)$ denotes the model energy conditioned on the target class label $\mathbf{y}^+$ which we specify as the negative logit corresponding to the target class label $\mathbf{y}^{+}$. To allow for faster sampling, we follow the common practice of choosing the step-size $\epsilon_j$ and the standard deviation of $\mathbf{r}_j$ separately. While $\mathbf{x}_J$ is only guaranteed to distribute as $p_{\theta}(\mathbf{x}|\mathbf{y}^{+})$ if $\epsilon \rightarrow 0$ and $J \rightarrow \infty$, the bias introduced for a small finite $\epsilon$ is negligible in practice \citep{murphy2023probabilistic,grathwohl2020your}. Appendix~\ref{app:jem} provides additional implementation details for any tasks related to energy-based modelling.

@@

Our default choice for the $\text{dist}(\cdot)$ function in both cases is the Euclidean Norm.

@@

\begin{equation} \label{eq:eccco}
  \begin{aligned}
  \mathbf{Z}^\prime &= \arg \min_{\mathbf{Z}^\prime \in \mathcal{Z}^L} \{  {\text{yloss}(M_{\theta}(f(\mathbf{Z}^\prime)),\mathbf{y}^+)}+ \lambda_{1} {\text{dist}(f(\mathbf{Z}^\prime),\mathbf{x}) } \\
  &+ \lambda_2 \Delta\mathcal{E}_{\theta}(\mathbf{Z}^\prime,\widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}) + \lambda_3 \Omega(C_{\theta}(f(\mathbf{Z}^\prime);\alpha)) \} 
  \end{aligned} 
\end{equation}

@@

\begin{equation} \label{eq:energy-delta}
  \begin{aligned}
    \Delta\mathcal{E}_{\theta}&=\mathcal{E}_{\theta}(f(\mathbf{Z}^\prime)|\mathbf{y}^+)-\mathcal{E}_{\theta}(x|\mathbf{y}^+) &&& x \sim \widehat{\mathbf{X}}_{\theta,\mathbf{y}^+}
  \end{aligned}
\end{equation}
